---
title: "Group Project Preparation"
author: "World Cup Finals Group"
date: "`r Sys.Date()`"
output: html_document
---

## World Cup Dataset 

### Introduction

The FIFA World Cup Dataset covers every single World Cup match played in its history from Uruguay in 1930 to Russia in 2018. The dataset consists of two spreadsheets. The `wcmatches.csv` file contains every football match played in the World Cup, with some columns like which year, which team claimed victory, and what the win condition was. The `worldcups.csv` file contains the winners and has some dataset features like winners, games, goals_scored and more.

### Data cleaning and summary

Firstly, we prepare the two datasets using the `readr` package.

```{r Reading Data}
library(readr)

root_url <- 'https://raw.githubusercontent.com/'
intermediate <- 'rfordatascience/tidytuesday/master/data/2022/2022-11-29/'

wcmatches <- read_csv(paste0(root_url, intermediate, 'wcmatches.csv'))
worldcups <- read_csv(paste0(root_url, intermediate, 'worldcups.csv'))
```

After that, we will do some cleaning with the two datasets. 
For the `worldcups` dataset, we first pivot the data longer so that we are able to condense the winner to fourth columns to a column called `position_status` to make it cleaner. After that, we change "USA" to "United States" to make it consistent with the `wcmatches` dataset, and then we add in a new feature called continent for our visualisation later.

For the `wcmatches` dataset, we rename some of the columns, namely `stage_status` from `stage` and `win_outcome` from `outcome.` After that, we mutate the `stage_status` such that if the datapoint consists of "Group", we convert it to "Group Stage" and if the datapoint equals "Final Round", we convert it to "Finals" to make it cleaner. We remove the date because we already have the month and year, and we expand the `win_outcome` column to make it more clear on what H, D and A are. After which, we calculate the score difference, which is the difference in the scoreline, and we fill the `win_conditions` column with Nil before extracting two more features from the `win_conditions` column, namely the `after_extra_time` and `penalties` column.

These are all in preparation to answer the question - what factors affect the performance of countries in the World Cup?

```{r Data Cleaning}
library(tidyverse)
library(stringr)

americas <- c("Uruguay", "Brazil", "Chile", "Mexico", "Argentina", "United States")

europe <- c("Italy", "France", "Switzerland", "Sweden", "England", "Germany", "Spain", "West Germany", "Netherlands", "Austria", "Belgium", "Croatia", "Czechoslovakia", "Hungary", "Poland", "Portugal", "Yugoslavia", "Bulgaria", "Soviet Union", "Turkey", "Russia")

others <- c("South Africa", "Japan", "South Korea", "Japan, South Korea")

worldcups_tidy <- worldcups %>%
                  pivot_longer(winner:fourth, names_to = "position_status", values_to = "country_name") %>%
                  rename(host_country = host) %>%
                  mutate(host_country = ifelse(host_country == "USA", "United States", host_country), 
                         country_name = ifelse(country_name == "USA", "United States", country_name)) %>% mutate(host_continent = case_when(
                         host_country %in% americas ~ "Americas", 
                         host_country %in% europe ~ "Europe",
                         host_country %in%  others ~ "Rest Of World"), 
                         country_continent = case_when(
                         country_name %in% americas ~ "Americas", 
                         country_name %in% europe ~ "Europe",
                         country_name %in%  others ~ "Rest Of World")) 

wcmatches_tidy <- wcmatches %>%
                  rename(stage_status = stage, win_outcome = outcome) %>%
                  mutate(stage_status = ifelse(str_detect(stage_status, "Group"), "Group Stage", stage_status), 
                         stage_status = ifelse(stage_status == "Final Round", "Finals", stage_status)) %>%
                  select(-date) %>%
                  mutate(win_outcome = case_when(
                         win_outcome == "H" ~ "Home",
                         win_outcome == "D" ~ "Draw",
                         win_outcome == "A" ~ "Away")) %>%
                  mutate(score_difference = abs(home_score - away_score), 
                         win_conditions = replace_na(win_conditions, "Nil"),
                         after_extra_time = ifelse(str_detect(win_conditions, "AET"), 1, 0), 
                         penalties = ifelse(str_detect(win_conditions, "penalties"), 1, 0)) 
```


### Visualisations 

1st Visualisation: 
**Create a bar plot showcasing the top ten countries in the world cup that have the best overall performance. Include all years**

**How to Calculate the best overall performance?** 

Point Allocation Should be: 

  Case 1: If Home Score > Away Score before extra time, the Home Team will be allocated 3 points while the Away Team will be allocated 0 points. 
  
  Case 2: If Away Score > Home Score before extra time, the Away Team will be allocated 3 points while the Home Team will be allocated 0 points.
  
  Case 3: If Home Score = Away Score before extra time and Home Team wins the match after extra time, the Home Team will be allocated 3 points while the Away Team will be allocated 0 points. 
  
  Case 4: If Home Score = Away Score before extra time and Away Team wins the match after extra time, the Away Team will be allocated 3 points while the Home Team will be allocated 0 points. 
  
  Case 5: If Home Score = Away Score before extra time and Home Score = Away Score after extra time, we treat the match as a draw and 1 point will be allocated to both the Home Team and the Away Team. 
  
Sum up all the points for each country in the world cup and plot the overall performance of the top 10 best countries in descending order using a bar graph. 
This involves data manipulation techniques and joining multiple datasets.

```{r}
library(ggplot2)
library(forcats)

# For Cases 1 and 3, where the Home Score > Away Score. In these cases, we select the Home Teams and count the number of times each country has appeared. Since 3 points will be allocated for each win, we should multiply the number of observations by 3 after the n() function. 

wcmatches_home_better <- wcmatches_tidy %>%
                         filter(home_score > away_score) %>%
                         select(home_team) %>%
                         group_by(home_team) %>%
                         summarize(total_home_points = n() * 3) %>%
                         rename(team = home_team)

# For Cases 2 and 4, where the Away Score > Home Score. In these cases, we select the Away Teams and count the number of times each country has appeared. Since 3 points will be allocated for each win, we should multiply the number of observations by 3 after the n() function.

wcmatches_away_better <- wcmatches_tidy %>%
                         filter(away_score > home_score) %>%
                         select(away_team) %>%
                         group_by(away_team) %>%
                         summarize(total_away_points = n() * 3) %>%
                         rename(team = away_team)

# For Case 5. These are the cases where Away Score = Home Score even after extra time. In this case, we need to select both the Home Team and the Away Team. I have split the draw matches into two sets, one set containing the Home Team countries and the other set containing the Away Team countries. Calculate the points for the draw matches accordingly. Since only 1 point will be allocated for each draw to each team, we just need to use n(). 

wcmatches_draw_games_1 <- wcmatches_tidy %>%
                          filter(home_score == away_score) %>%
                          select(home_team) %>%
                          group_by(home_team) %>%
                          summarize(home_draw_points = n())

wcmatches_draw_games_2 <- wcmatches_tidy %>%
                          filter(home_score == away_score) %>%
                          select(away_team) %>%
                          group_by(away_team) %>%
                          summarize(away_draw_points = n())

# Join the two datasets containing the Home Teams and Away Teams respectively to obtain a dataset that contains the points for each scenerio (where the country is acting as a Home Team and Away Team). We need to use a full_join to join the datasets as we do not want any loss of observations from either dataset. Those observations that are not in one dataset will be denoted by NA when merged, hence need to change the value from NA to 0. Next, create a new variable that sums up the points for the country acting as Home Team and Away Team. 

wcmatches_draw <- wcmatches_draw_games_1 %>%
                  full_join(wcmatches_draw_games_2, by = c("home_team" = "away_team")) %>%
                  mutate(home_draw_points = replace_na(home_draw_points, 0),
                         away_draw_points = replace_na(away_draw_points, 0),
                         total_draw_points = home_draw_points + away_draw_points) %>%
                  rename(team = home_team) %>%
                  select(team, total_draw_points)

# Afterwards, join the three datasets (points for Home Team Win, points for Away Team Win and points for Draw Matches), again using a full_join to prevent any loss of observations. Similarly, observations that are not in one or more datasets will be denoted by NA when merged, hence need to change the value from NA to 0. Next, create a new variable that sums up the home points, away points and draw points. Use head(10) to filter the top 10 best performing countries after applying the arrange function in descending order of the final score.  

wcmatches_tabulated <- wcmatches_home_better %>%
                       full_join(wcmatches_away_better, by = "team") %>%
                       full_join(wcmatches_draw, by = "team") %>%
                       mutate(total_home_points = replace_na(total_home_points, 0),
                              total_away_points = replace_na(total_away_points, 0),
                              total_draw_points = replace_na(total_draw_points, 0), 
                              final_score = total_home_points + total_away_points + total_draw_points) %>% 
                       select(team, final_score) %>%
                       arrange(desc(final_score)) %>%
                       head(10)

# Create a bar plot

ggplot(wcmatches_tabulated, aes(x = fct_reorder(team, final_score, .desc = TRUE), y = final_score)) + 
  geom_col(fill = "red") + 
  labs(x = "Country", y = "Grand Total Score", title = "Top 10 Best Performing Countries In World Cup") + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  geom_text(label = wcmatches_tabulated$final_score, size = 5, position = position_stack(vjust = 0.5)) + 
  theme(axis.text.x = element_text(size = 7))
```


Performance based on rankings in each world cup (work in progress, I need to think of a good way to do this)
- average ranking across all world cups attended?
- categories (winner, 2nd, 3rd, 4th, top8, top16, group stage)?
```{r}

```

Host country performance
```{r}

```



### DO NOT DELETE THIS
1. find country with best overall performance --> hypothesize with the next few questions
2. line plot
3. scatterplot 
